Bluetoothを用いたマルチVM対応\\mrubyバイトコードローダ

### 2016-02-17 ETNET2016/thesis.tex
add ETNET2016

%
% File: thesis.tex
% Created: 月 11 30 08:00 午後 2015 東京 (標準時)
% Last Change: 月 11 30 08:00 午後 2015 東京 (標準時)
%
\input{dummy/dummy.tex}
\documentclass[conference,compsoc]{IEEEtran}
\usepackage[dvipdfmx]{graphicx}
\usepackage{amssymb}
\usepackage{enumerate,cite,url}
\usepackage{listings,jlisting}
\lstset{%
    language={c},%
    basicstyle={\small},%
    identifierstyle={\small},%
    commentstyle={\small\itshape},%
    keywordstyle={\small},%\bfseries},%
    ndkeywordstyle={\small},%
    stringstyle={\small\it},
    frame={tb},
    breaklines=true,
    columns=[l]{fullflexible},%
    numbers=left,%
    xrightmargin=0zw,%
    xleftmargin=3zw,%
    numberstyle={\scriptsize},%
    stepnumber=1,
    numbersep=1zw,%
    %lineskip=-0.5ex%
}

\title{mruby Bytecode Loader Using Bluetooth in Multi-VM Environment}
\author{Takuro Yamamoto}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
In recent years, the productivity of embedded systems has become a problem due to their complexity and large-scale.
For the purpose of improving the productivity for embedded software development, the mruby on TECS framework has been proposed that is applied mruby (Lightweight Ruby) and supports component-based development.
In the current mruby on TECS, the mruby programs have to be compiled and linked every time the programs are modified because the mruby bytecodes are incorporated in the platform.
Moreover, while the framework supports multi-VM, developers need to be familiar with the functions of real-time operating systems to effectively execute multiple mruby programs in concurrent or/and parallel.
To improve the development efficiency, this paper proposes an mruby bytecode loader using Bluetooth as an extension of mruby on TECS.
The loader executes two mruby bytecodes, mruby application bytecode and mruby library bytecode.
mruby application bytecode modified frequently is sent from a host to a target device by developers.
mruby library bytecode modified infrequently is preserved beforehand with the platform in a strage device at the time of the first compilation.
In addition, multiple mruby programs cooperatively run in the proposed framework.
A RiteVM sceduler makes multitasking processing more easy-to-use than that of mruby on TECS.
Synchronization of starting multiple tasks is also implemented with Eventflag. 
Experimental results demonstrate the advantages of the proposed framework.
\end{abstract}
\section{Introduction}
In these days, embedded systems have been required the high-quality and the high-performance.
Due to this trend, the complexity of embedded systems also increases and the scale is larger.
For example, IoT (Internet of Things) applications.
The low production cost and the short developing period of time are also needed.

An approach of efficient software development is to use component-based techniques.
CBD (Component-Based Development) is a design technique for constituting reusable components \cite{par:Crnkovic}.
Complex and large scale software systems can be developed efficiently using component-based techniques.
That is because software componentization provides high-reusability and easy verification.
It also makes a system flexible in extensions and specification changes.
There are TECS \cite{par:TECS}, AUTOSAR \cite{url:AUTOSAR}, and SaveCCM \cite{par:SAVEapproach} as a typical component-based development for embedded systems

In addition, another approach of efficient software development is to develop with script languages.
Most of current software are programmed in C language, and the development with C takes a high cost and more time to develop.
Script languages make software engineering more efficient and shorten a development period because script languages have high-productivity from their usability.
Java script, Perl, Python, Lua and Ruby are well-known as representative script languages.

Although script languages are easy to use and read, their execution time are slower than C language's.
For embedded systems, the real-time properties such as worst-execution time and response time are very important factors.
Therefore, it is difficult to apply the script languages to embedded systems.

mruby on TECS is a component-based framework for running script program \cite{par:mrubyonTECS}.
It is integrated two technologies.
One is mruby, which is a script language for embedded systems \cite{par:mruby}, \cite{url:mruby}.
The other is TECS (TOPPERS Embedded Component System), which is a component-based framework for embedded systems \cite{par:TECS} \cite{url:TOPPERS}.
mruby on TECS supports to effectively run mruby script language on embedded systems.
mruby on TECS also makes execution time 100 times faster than that of mruby.

%This paper proposes two additional features of mruby on TECS, mruby bytecode loader using Bluetooth and RiteVM sceduler.
mruby on TECS has several problems at present.
One of the problems is low software development efficiency.
mruby on TECS only supports a storage/ROM device in the platform for LEGO MINDSTORMS EV3 \cite{par:EV3} to load mruby programs.
For example, an SD card should be inserted and pulled out repeatedly, or a ROM should be rewritten if mruby programs are modified.
Developers should also restart an OS on a target device.
%Moreover, in the current system if developers attempt to run multiple tasks with the same priority, a task with first execution only runs and all other tasks do not run unless developers call the OS's function.
%In addition developers must prepare as many VMs as the number of tasks because one RiteVM supports only one task.
Moreover, although mruby on TECS has supported the multi-VM, developers need to call OS's function in order to execute multiple tasks.
These problem are a big burden on developers.

This paper proposes an extended framework of mruby on TECS: an mruby bytecode loader using Bluetooth and a RiteVM scheduler for fairly executing mruby programs.
To improve the development efficiency, the mruby bytecode loader using Bluetooth enables developers to write the platform in a strage device only once at the beginning and to transfer mruby application codes from a host to a target device.
The RiteVM scheuler enables developers to run multitasking programs more easily than mruby on TECS.
%Specifically, a cyclic handler of switching tasks is implemented.
%This paper evaluates the proposed framework in term of data transfer rate, overhead of cyclic handler and execution time.

{\bf Contribution}: The proposed framework gives the contribution in the following points.
\begin{enumerate}
    \item {\bf To improve the software development efficiency:} Developers do not need to rewrite a storage/ROM device and also to restart an OS.
The loader supports the countinuous loading, which saves the Bluetooth set-up time.
Therefore, the mruby bytecode loader using Bluetooth helps developers develop software.
\item {\bf To effectively execute multiple mruby programs in concurrent or/and parallel:} Developers can execute multiple tasks without the knowledge of an RTOS because the RiteVM sceduler switches tasks cyclically.
In additon, synchronization of multipel tasks is also implemented.
\item {\bf To focus on the benefits of component-based developments:} This paper shows the specific examples for the benefits of component-based development. 
\end{enumerate}

{\bf Organization}: The paper is organized as follows.
Section \ref{sec:Background} introduces the basic technologies i.e. mruby, TECS and mruby on TECS.
Section \ref{sec:Design and Implementation} describes the design and implementation of the proposed framework in detail.
Section \ref{sec:Evaluation} evaluates the proposed framework, Section \ref{sec:Related work} discusses related work, and then Section \ref{sec:Conclusion} concludes this paper.

\section{Background}
\label{sec:Background}
Figure \ref{fig:proposed} shows the system model of the proposed framework.
The bytecodes are transferred from the host to the target device with Bluetooth.
The RiteVMs and mruby library are assumed to be prepared in advance.
Each bytecode is allocated to a RiteVM, respectively.
Developers can run bytecodes transferred with Bluetooth in multitask.

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/proposed.pdf}
    \caption{System Model}
    \label{fig:proposed}
\end{figure}

This section describes mruby on TECS on which the proposed framework is based.
mruby on TECS is a component-based framework for running script programs.
In mruby on TECS, two technologies are integrated: mruby and TECS.
To explain the system, mruby and TECS are also respectively described in this section.

\subsection{mruby}
\label{sec:mruby}
mruby is the light-weight implementation of Ruby programming language complying to part of the ISO standard.

Ruby is a object-oriented script language \cite{url:Ruby}.
As the main feature, Ruby is easy-to-use and easy-to-read due to its simple grammar.
Ruby can run a program with fewer code lines than C language.
Ruby improves the productivity of a software development owing to not only simple grammar also object-oriented functions such as classes and methods, exceptions, and garbage collection.

mruby is suitable for embedded systems because of faster execution with less amount of resources and takes over the usability and readability of Ruby.
In addition, VM (Virtual Machine) mechanism is used in mruby, therefore mruby programs can run on any operating system as long as VM is implemented.
A RiteVM is a VM in mruby, that runs mruby programs.
The RiteVM mechanism is shown in Figure \ref{fig:mruby}.
The mruby compiler translates an mruby code into a bytecode, which is an intermediate code that can be interpreted by a RiteVM.
The bytecodes can run on a Rite VM, and thus mruby programs can be executed on any target devices if only RiteVMs are implemented.
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/mruby.pdf}
    \caption{Mechanism of mruby/RiteVM}
    \label{fig:mruby}
\end{figure}

\subsection{TECS}
\label{sec:TECS}
TECS (TOPPERS Embedded Component System) is a component system suitable for embedded systems.
TECS helps decrease complexity and difficulty because the generated component diagram can visualize the structure of whole software.
It can also help increase productivity and reduce development duplication  because a common part of software is regarded as a component.

The component deployment and composition in TECS are statically performed, which gives optimization.
As a result, the overhead of execution time and memory consumption can be reduced.
There are other features of TECS, implementation in C language, source level portability, fine-grained component, etc.

\subsubsection{Component Model}\mbox{}\\

Figure \ref{fig:component} shows a component diagram.
A {\it cell} which is an instance of a component in TECS consists of {\it entry} ports, {\it call} ports, attributes and internal variables.
A {\it entry} port is an interface to provide functions with other {\it cell}s, and a {\it call} port is an interface to use functions of other {\it cell}s.
A {\it cell} has one or more {\it entry} ports and {\it call} ports.
Functions of a {\it cell} are implemented in the C language.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/component_diagram.pdf}
    \caption{Component Diagram}
    \label{fig:component}
\end{figure}

\subsubsection{Component Description}\mbox{}\\

The description of a component in TECS is divided into three parts, {\it signature}, {\it celltype}, and build description.
TECS code is written in .cdl (component description language) file.
The component description is mentioned with an example shown in Figure \ref{fig:component} as follow.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/mrubyontecs.pdf}
    \caption{System Model of existing mruby on TECS}
    \label{fig:mrubyontecs}
\end{figure}

@@

Figure \ref{fig:mruby_TECS_bridge} shows an example of use of an mruby-TECS bridge for controlling a motor.
The left side of BridgeMotor belongs to the mruby program.
The right side of BridgeMotor belongs to TECS component.
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/mruby_TECS_bridge.pdf}
    \caption{mruby-TECS Bridge}
    \label{fig:mruby_TECS_bridge}
\end{figure}

@@

\section{Design and Implementation}
\label{sec:Design and Implementation}
Figure \ref{fig:system_model} shows the detailed system model of the proposed framework.
Each mruby application bytecode transferred from the host is received by the loader in the RiteVM.
The RiteVM reads the own bytecode and executes it.
The mruby applications run at the same time because of synchronization processing.
The RiteVM scheuler switches tasks as multiple tasks can run in concurrent.

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/system_model.pdf}
    \caption{Detailed System Model of the proposed framework}
    \label{fig:system_model}
\end{figure}

\subsection{mruby Bytecode Loader Using Bluetooth}
\label{sec:mruby bytecode loader using Bluetooth}
This section describes an additional functionality of mruby on TECS, mruby bytecode loader using Bluetooth.
In the current system, the platform including mruby bytecodes are saved in a storage/ROM device.
Developers must rewrite the storage/ROM device every time the application programs are modified.
The OS on the target device should be also restarted.
It causes a low development efficiency to repeat that.
An mruby bytecode loader using Bluetooth makes the developers' burden decrease.
%The work that an SD card is in and out needs to be done just once in the beginning.
Developers should just once carry out the work to connect the storage/ROM device and start the OS.

mruby programs are consisted of mruby application and libraries.
mruby application is the main code, which developers should program.
mruby libraries are the codes defining the functions for application such as Ruby classes. 
The mruby bytecodes including mruby application and libraries can be sent and run.
However, it is also wasteful in terms of the size and time to send because libraries are not frequently modified. 
The proposed framework provides the design that only mruby applications are sent and mruby libraries are preserved with the platform in the storage/ROM device beforehand.
Due to this design, RiteVMs can share the mruby libraries.
In additon, a RiteVM can use the own library that other RiteVMs should not use.

%The development flow in an mruby bytecode loader using Bluetooth is shown in Figure \ref{fig:bluetooth_loader}.
In the proposed framework, the platform including RiteVMs and mruby library is compiled and copied in the storage/ROM device at the first.
%These are assumed not to be modified.
%The binary data transferred with Bluetooth is the bytecode of the main source.
In the host, the mruby application programs (.rb) are edited and compiled into the bytecodes (.mrb) by an mruby compiler.
The generated bytecodes are transferred from the host to the target device with Bluetooth.

Moreover, developers can save the time of Bluetooth pairing since the loader can continuously load the bytecode.

% \begin{figure}[t]
%     \centering
%     \includegraphics[width=8cm,clip]{figure/bluetooth_loader.pdf}
%     \caption{Development Flow in mruby bytecode loader using Bluetooth}
%     \label{fig:bluetooth_loader}
% \end{figure}

\subsubsection{Component of RiteVM with mruby bytecode loader using Bluetooth} \mbox{}\\

The proposed framework provides a RiteVM with mruby bytecode loader using Bluetooth as TECS component.
The component is an extension of the RiteVM component, which is described in \cite{par:mrubyonTECS}.
The component plays a role in receiving bytecodes via Bluetooth, and also manages a RiteVM configuration such as automatically generates the bytecode in the build description.
This generated bytecode is prepared beforehand in the storage/ROM device such as mruby libraries, and different from a bytecode transferred with Bluetooth.

Figure \ref{fig:component_bluetooth} shows a component diagram of MrubyTask1 and MrubyBluetooth1 {\it cell}s.
The MrubyTask1 {\it cell} is a componentized task of the RTOS (TOPPERS/HRP2).
TOPPERS/HRP2 is described in \cite{url:HRP2}, \cite{par:hr-tecs}.
The MrubyBluetooth1 is a component of RiteVM with mruby bytecode loader using Bluetooth.
A bytecode in the host is transferred and received at the top of the component.
In this framework, ZMODEM \cite{par:zmodem} is used as a binary transfer protocol.

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{figure/component_bluetooth.pdf}
    \caption{Component Diagram of mruby bytecode loader using Bluetooth}
    \label{fig:component_bluetooth}
\end{figure}

Figure \ref{fig:control_flow} shows the process of executing mruby program in a component of RiteVM with mruby bytecode loader using Bluetooth such as MrubyBluetooth1.
The concrete main code of tMrubyBluetooth is shown in Figure \ref{maincode_mrubybluetooth}.

First, a pointer of {\it mrb\_state} and {\it mrbc\_context} are initialized.
{\it mrb\_state} is a set of states and global variables used in mruby.

Second, the RiteVM reads the bytecode of mruby libraries.
mruby libraries are a set of Ruby classes such as motor class and sensor class.
For example, motor class defines methods to rotate and stop a motor.
The tMrubyBluetooth {\it cell} has attributes as shown in Figure \ref{celltype_mrubybluetooth}.
The {\it mrubyFile} indicates the program files of mruby libraries.
{\it [omit]} is only used for the TECS generator, thus the attribute, {\it mrubyFile}, does not consume memory.
The {\it irep} is the pointer of the array stored the bytecode of mruby libraries.
In short, the bytecode of mruby libraries is stored as an attribute of the component when compiling for the first time.

Third, the RiteVM reads the bytecode of the mruby application transferred with Bluetooth.
The mruby application bytecode is stored in an array of type uint8\_t such as {\it irepApp}, which is a tMrubyBluetooth variable as shown in \ref{celltype_mrubybluetooth}.
The array is different from that of holding the mruby library bytecode.
Two bytecodes are read separately in the RiteVM.

>>>>

%        File: thesis.tex
%     Created: Tue Feb 16 06:00 PM 2016 J
% Last Change: Tue Feb 16 06:00 PM 2016 J
%
\input{../dummy/dummy.tex}
\documentclass[submit]{ipsj_v2/UTF8/ipsj}
%\documentclass{ipsj}

\usepackage{latexsym}
\usepackage[dvipdfmx]{graphicx}
\usepackage{amssymb}
\usepackage{enumerate,cite,url}
\usepackage{listings,jlisting}
\lstset{%
    language={c},%
    basicstyle={\small},%
    identifierstyle={\small},%
    commentstyle={\small\itshape},%
    keywordstyle={\small},%\bfseries},%
    ndkeywordstyle={\small},%
    stringstyle={\small\it},
    frame={tb},
    breaklines=true,
    columns=[l]{fullflexible},%
    numbers=left,%
    xrightmargin=0zw,%
    xleftmargin=3zw,%
    numberstyle={\scriptsize},%
    stepnumber=1,
    numbersep=1zw,%
    %lineskip=-0.5ex%
}

\def\Underline{\setbox0\hbox\bgroup\let\\\endUnderline}
\def\endUnderline{\vphantom{y}\egroup\smash{\underline{\box0}}\\}
\def\|{\verb|}

\setcounter{巻数}{57}
\setcounter{号数}{1}
\setcounter{page}{1}

\受付{2015}{3}{4}
%\再受付{2015}{7}{16}   %省略可能
%\再再受付{2015}{7}{20} %省略可能
\採録{2015}{8}{1}

\begin{document}

\title{Bluetoothを用いたマルチVM対応\\mrubyバイトコードローダ}

\etitle{mruby Bytecode Loader Using Bluetooth in Multi-VM Environment}

\affiliate{IPSJ}{情報処理学会\\
IPSJ, Chiyoda, Tokyo 101--0062, Japan}

\paffiliate{JU}{情報処理大学\\
Johoshori University}

\author{山本 拓朗}{Takuro Yamamoto}{IPSJ}[joho.taro@ipsj.or.jp]
\author{大山 博司}{Hiroshi Oyama}{IPSJ}
\author{安積 卓也}{Takuya Azumi}{IPSJ,JU}[gakkai.jiro@ipsj.or.jp]

\begin{abstract}
近年，組込みシステムは複雑化・大規模化しているため，組込みソフトウェアの生産性が問題になっている．

組込みソフトウェア開発の生産性の向上を目的として，mruby（軽量Ruby）を適用させたコンポーネントベース開発が可能なフレームワークであるmruby on TECSを提案してきた．

現状のmruby on TECSでは，プラットフォームにmrubyバイトコードを組み込んでいるため，mrubyプログラムを修正する度にコンパイル・リンクし直す必要がある．
さらに，マルチVMを提供しているが，複数のmrubyプログラムを効率よく並行動作させるには開発者がリアルタイムOSの機能を熟知している必要がある．

本研究では，mruby on TECSの拡張として，mrubyアプリケーションのバイトコードをBluetoothで転送することで開発効率を向上させる．
さらに，複数のmrubyプログラムを協調動作できるフレームワークを提案する.

\end{abstract}

\begin{jkeyword}
情報処理学会論文誌ジャーナル，\LaTeX，スタイルファイル，べからず集
\end{jkeyword}

\begin{eabstract}
This document is a guide to prepare a draft for submitting to IPSJ
Journal, and the final camera-ready manuscript of a paper to appear in
IPSJ Journal, using {\LaTeX} and special style files.  Since this
document itself is produced with the style files, it will help you to
refer its source file which is distributed with the style files.
\end{eabstract}

\begin{ekeyword}
IPSJ Journal, \LaTeX, style files, ``Dos and Don'ts'' list
\end{ekeyword}

\maketitle

\section{はじめに}
近年，組込みシステムは高品質・高性能化に伴い，複雑化・大規模化している．
例えば，IoTアプリケーションがある．
その上，製品の低コスト化や短期間での開発も要求されている．

ソフトウェアを効率的に開発するアプローチのひとつに，コンポーネントベース開発がある．
コンポーネントベース開発は，再利用可能なコンポーネントを構成するための設計手法である\cite{par:Crnkovic}．
コンポーネント化されたソフトウェアは，再利用性が高く，検証が容易になるため，コンポーネントベース開発を適用することで，複雑かつ大規模なソフトウェアを効率的に開発することができる．
システムの拡張や仕様変更にも柔軟に対応できる．
組込みシステムのコンポーネント技術として，TECS\cite{par:TECS}やAUTOSAR\cite{url:AUTOSAR}，SaveCCM\cite{par:SAVEapproach}がある．

効率的なソフトウェア開発のもう一つのアプローチとして，スクリプト言語がある．
現在のソフトウェアのほとんどはC言語で開発されているが，C言語による開発はコストが高く，開発期間も長くなる．
スクリプト言語は，その使いやすさから高い生産性をもっているため，効率的かつ短期間での開発ができる．
有名なスクリプト言語として，Java script，Perl，Python，Lua，Rubyなどがある．

しかし，スクリプト言語は可読性や使いやすさが高い反面，C言語よりも実行時間が遅い．
組込みシステムにとって，最悪実行時間や応答時間といったリアルタイム性は非常に重要であるため，スクリプト言語を組込みシステムに利用することは難しい．

%mruby on TECSは，スクリプト言語であるmruby使用したコンポーネントベース開発が可能なフレームワークである\cite{par:mrubyonTECS}．
mruby on TECS は，mruby(軽量Ruby) と，組込みシステムに適したコンポーネントシステムであるTECS (TOPPERS Embedded Component System) を組み合わせたフレームワークである\cite{par:mrubyonTECS}.
mruby on TECS では，mruby プログラムからC 言語の関数を呼ぶ機能を提供しており，mruby に比べて，アプリケーションを約100 倍速く実行できる.

現状のmruby on TECSでは，いくつかの問題がある．
mruby on TECS は，mrubyプログラムをロードするための方法として，記憶装置しか対応していないため，作業効率が悪い．
(LEGO MINDSTORMS EV3 のプラットフォーム\cite{par:EV3})
mrubyプログラムを修正する度に，SDカードの抜き差しやROMへの書き込みを行う必要がある．
%mruby on TECS では，mruby プログラムを修正する度にターゲットデバイスでSD カードまたはROMに書き込みし，OS を再起動する必要がある.
さらに，mruby on TECS はマルチVMに対応してるが，複数のmruby アプリケーションを並行実行させる場合，開発者がタスクを待ち状態へ遷移させるOS の機能を呼び出さなければならない.

% mrubyバイトコードローダによって，開発者ははじめの一度だけ記憶装置にプラットフォーム部分を書き込み，mrubyアプリケーションのバイトコードをホストからターゲットデバイスへBluetoothを用いて転送し，実行することが可能になる．
% 繰り返してSDカードの抜き差しやROMへの書き込みをする手間が省けるため，作業効率は向上する．
% RiteVMスケジューラは複数のmrubyアプリケーションを
提案フレームワークは，mruby on TECS を拡張して，Bluetooth を用いたmruby バイトコードローダと，実用的なマルチタスク処理の実装を行った.
提案フレームワークでは，プラットフォーム部分をコンパイル・リンクし，ターゲットデバイス上で起動する.
ホストPC 側では，mruby アプリケーション(.rb) をバイトコード(.mrb) にコンパイルし，Bluetooth を通してターゲットデバイスにバイトコードを転送する. 
ターゲットデバイス側では，RiteVM に実装されたローダが転送されたバイトコードを受信し，すでにプラットフォームに含まれているmruby ライブラリと合わせてアプリケーションを実行する.
これによって，繰り返しSDカードの抜き差しやROMへの書き込みをする手間やOSを再起動する時間が省けるため，作業効率を上げることができる.
さらに，各VM の処理を平等に実行するRiteVM スケジューラを提供することで，マルチタスク処理を効率的に利用できる.

% Contribution: The proposed framework gives the contribution in the following points.
% 1) To improve the software development efficiency:
% Developers do not need to rewrite a storage/ROM device and also to restart an OS. 
% The loader supports the countinuous loading, which saves the Bluetooth setup time. 
% Therefore, the mruby bytecode loader using Bluetooth helps developers develop software.
% 2) To effectively execute multiple mruby programs in concurrent or/and parallel: 
% Developers can execute multiple tasks without the knowledge of an RTOS because the RiteVM sceduler switches tasks cyclically.
% In additon, synchronization of multiple tasks is also implemented.
% 3) To focus on the benefits of component-based developments:
% This paper shows the specific examples for the benefits of component-based development.

% Organization: The paper is organized as follows. 
% Section 2 introduces the basic technologies i.e. mruby, TECS and mruby on TECS. 
% Section 3 describes the design and implementation of the proposed framework in detail. 
% Section 4 evaluates the proposed framework, Section 5 discusses related work, and then Section 6 concludes this paper.

\section{既存フレームワーク}
\label{fig:background}
図\ref{fig:proposed}に提案フレームワークのシステムモデルを示す．
RiteVMとmrubyライブラリを含むプラットフォーム部分は，はじめにターゲットデバイス上で起動されている．
mrubyアプリケーションのバイトコードは，ホストからターゲットデバイスへと転送される．
それぞれバイトコードは，RiteVMに割り当てられ，並行して起動される．

この章では，提案フレームワークのベースになっているmruby on TECSについて述べる．
mruby on TECSで利用されているmrubyとTECSについても述べる．

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/proposed.pdf}
    \caption{System Model}
    \label{fig:proposed}
\end{figure}

\subsection{mruby}
mrubyは，ISO規格の一部に準拠したRubyの軽量実装である．
Ruby\cite{url:Ruby}は，オブジェクト指向型のスクリプト言語である．
特徴としてシンプルな文法なため読みやすくかつ使いやすい上に，C言語よりも少ないコード量で記述できる．
他にも，クラスやメソッドといったオブジェクト指向関数，例外，ガベージコレクションなどの機能により，生産性を向上させることができる．

mrubyは，Rubyの可読性や使いやすさはそのままに，リソースが少なく起動が速いため，組込みシステムに適している．
mrubyには，RiteVMと呼ばれるVMが実装されているため，どのOS上でもアプリケーションを起動させることができる．
図\ref{fig:mruby}にRiteVMのメカニズムを示す．
mrubyコンパイラは，mrubyプログラム(.rb)を，中間言語であるバイトコード(.mrb)にコンパイルする．
RiteVMが実装されいれば，どのターゲットデバイス上でも同じmrubyプログラムを実行できる．

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/mruby.pdf}
    \caption{Mechanism of mruby/RiteVM}
    \label{fig:mruby}
\end{figure}

\subsection{TECS}
TECS (TOPPERS Embedded Component System)は，組込みシステム向けのコンポーネントシステムである．
TECSによるコンポーネントベース開発は，システム全体の構造を可視化するため，システムの複雑さや難しさを減らすことができる．
さらに，ソフトウェアの共通部分はコンポーネントとして扱われるため，開発の重複を減らし，生産性を向上させる．

TECSでのコンポーネントの生成と結合は静的に行われるため，最適化され，実行時間や消費メモリのオーバヘッドは少ない．
他の特徴として，C言語での実装，ソースレベルでの移植性，コンポーネントの粒度が小さいことなどがある．

\subsubsection{Component Model}
図\ref{fig:component}にコンポーネント図の例を示す．
{\it cell} which is an instance of a component in TECS consists of {\it entry} ports, {\it call} ports, attributes and internal variables.

A {\it entry} port is an interface to provide functions with other {\it cell}s, and a {\it call} port is an interface to use functions of other {\it cell}s.
A {\it cell} has one or more {\it entry} ports and {\it call} ports.
Functions of a {\it cell} are implemented in the C language.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/component_diagram.pdf}
    \caption{Component Diagram}
    \label{fig:component}
\end{figure}

\subsubsection{Component Description}
The description of a component in TECS is divided into three parts, {\it signature}, {\it celltype}, and build description.
TECS code is written in .cdl (component description language) file.
The component description is mentioned with an example shown in Figure \ref{fig:component} as follow.

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/mrubyontecs.pdf}
    \caption{System Model of existing mruby on TECS}
    \label{fig:mrubyontecs}
\end{figure}

@@

Figure \ref{fig:mruby_TECS_bridge} shows an example of use of an mruby-TECS bridge for controlling a motor.
The left side of BridgeMotor belongs to the mruby program.
The right side of BridgeMotor belongs to TECS component.
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/mruby_TECS_bridge.pdf}
    \caption{mruby-TECS Bridge}
    \label{fig:mruby_TECS_bridge}
\end{figure}

@@

\section{Design and Implementation}
\label{sec:Design and Implementation}
図\ref{fig:system_model}に提案フレームワークの詳細なシステムモデルを示す．
ホストから転送されたmrubyアプリケーションのバイトコードは，それぞれのRiteVMに実装されたローダで受信され，同期処理により同時に実行される．
RiteVMスケジューラがタスクを周期的に切り替えるため，mrubyアプリケーションは並行動作することができる．
　
Figure \ref{fig:system_model} shows the detailed system model of the proposed framework.
Each mruby application bytecode transferred from the host is received by the loader in the RiteVM.
The RiteVM reads the own bytecode and executes it.
The mruby applications run at the same time because of synchronization processing.
The RiteVM scheuler switches tasks as multiple tasks can run in concurrent.

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/system_model.pdf}
    \caption{Detailed System Model of the proposed framework}
    \label{fig:system_model}
\end{figure}

\subsection{mruby Bytecode Loader Using Bluetooth}
\label{sec:mruby bytecode loader using Bluetooth}
この章では，Bluetoothを用いたmrubyバイトコードローダについて述べる．
現状のmruby on TECSでは，mrubyバイトコードをプラットフォームに組み込んでいるため，mrubyプログラムを修正する度にプラットフォーム部分をもう一度コンパイル・リンクし，SDカードやROMといったデバイスに再度書き込み，OSを再起動する必要がある．
このような作業を繰り返すことで，作業効率は悪くなる．
提案フレームワークでは，プラットフォーム部分をコンパイル・リンクし記憶デバイスに書き込むのは，はじめの一度だけで良いため，作業効率を向上させることができる．

mrubyプログラムは，アプリケーション部分とライブラリ部分に分けられる．
mrubyアプリケーションは，開発者が主にプログラムするメインのコードであり，mrubyライブラリには，motorクラスやsensorクラスといったRubyクラスのようにアプリケーションで利用される関数群が定義されている．
アプリケーションとライブラリを含んだmrubyバイトコードを転送し，実行することもできるが，提案フレームワークでは，ライブラリ部分はプラットフォームに組込み，アプリケーション部分のみを転送する設計にした．
このような設計にすることで，ライブラリは頻繁に変更されないため，毎回コンパイルする無駄を省くことができる．
さらに，各RiteVMでライブラリを共有することや，RiteVM固有のライブラリを持つことも可能になる．

提案フレームワークでは，mrubyバイトコードの連続ローディングにも対応している．

\subsubsection{Component of RiteVM with mruby bytecode loader using Bluetooth}
提案フレームワークでは，mrubyバイトコードローダはTECSのコンポーネントとして提供されている．
このコンポーネントは，RiteVMのコンポーネントの拡張として実装されている．(詳細については\cite{par:mrubyonTECS}を参照)
このコンポーネントでは，転送されたバイトコードの受信に加え，RiteVMのコンフィグレーションが行われる．

図\ref{fig:component_bluetooth}に，MrubyTask1とMrubyBluetooth1のコンポーネントの例を示す．
MrubyTask1は，コンポーネント化されたRTOS (TOPPERS/HRP2 \ref{url:HRP2}\ref{par:HRP2})のタスクである．
MrubyBluetooth1は，mrubyバイトコードを実装したRiteVMのコンポーネントであり，このコンポーネントのはじめで転送されてきたバイトコードを受信する．
提案フレームワークでは，バイナリ転送プロトコルとして，ZMODEを適用した．

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/component_bluetooth.pdf}
    \caption{Component Diagram of mruby bytecode loader using Bluetooth}
    \label{fig:component_bluetooth}
\end{figure}

図\ref{fig:control_flow}に，ローダを実装したRiteVMがmrubyプログラムを実行するまでの制御フロー，図\ref{maincode_mrubybluetooth}には，RiteVMコンポーネントのソースコードを示す．

はじめに，{\it mrb\_state}と{\it mrbc\_context}のポインタを初期化する．
{\it mrb\_state}は，mrubyで使われる状態と変数のセットである．

次に，mrubyライブラリのバイトコードを読み込む．
図\ref{celltype_mrubybluetooth}に示すように，tMrubyBluetoothのセルは属性を持っている．
{\it mrubyFile}はmrubyライブラリのファイルを示しており，{\it [omit]}はTECSジェネレータによってのみ使えわれるため，この属性はメモリを消費することない．
{\it irep}は，mrubyライブラリのバイトコードが格納されている配列へのポインタである．
つまり，mrubyライブラリは始めのコンパイル時にコンポーネントの属性として，配列に保存されている．

次に，転送されてきたmrubyアプリケーションのバイトコードを読み込む．
mrubyアプリケーションのバイトコードは，図\ref{celltype_mrubybluetooth}に示す，内部変数uint8\_t型の配列{\it irepApp}に格納されている．
２つのバイトコードは，それぞれ別の配列に保存されており，別々に読み込まれる．

最後に，RiteVMはmrubyアプリケーションであるタスクを実行する．
mrubyアプリケーションのプログラムが修正された場合は，そのバイトコードのみを再転送する．

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/control_flow.pdf}
    \caption{Control Flow of mruby bytecode loader using Bluetooth}
    \label{fig:control_flow}
\end{figure}
\begin{figure}[t]
\centering
\begin{lstlisting}
celltype tMrubyBluetooth{
    entry sTaskBody eMrubyBody;
    attr{
      [omit]char_t *mrubyFile;
      char_t *irep = C_EXP("&$cell_global$_irep");
      uint32_t irepAppSize = C_EXP( BUFFER_SIZE );
    };
    var{
        [size_is(irepAppSize)] uint8_t *irepApp;
    };
};
\end{lstlisting}
\caption{Celltype Description for RiteVM with mruby bytecode loader using Bluetooth}
\label{celltype_mrubybluetooth}
\end{figure}
\begin{figure}[t]
\centering
\begin{lstlisting}
void
eMrubyBody_main( CELLIDX idx )
{
  /* Declaration variables */
  mrb_state *mrb;
  mrbc_context *c;
  /* New interpreter instance */
  mrb = mrb_open();
  // Omit: error check for mrb_state
  /* New mruby context */
  c = mrbc_context_new( mrb );
  // Omit: initialization of mruby-TECS bridge
  /* Receive the bytecode via Bluetooth */
  bluetooth_loader( VAR_irepApp );
  /* Load mruby library bytecode and run */
  mrb_load_irep_cxt( mrb, ATTR_irep, c );
  /* Load mruby application bytecode and run */
  mrb_load_irep_cxt( mrb, VAR_irepApp, c );
  if ( mrb->exc ) {
    /* Failure to execute */
    mrb_p( mrb, mrb_obj_value( mrb->exc ) );
    exit( 0 );
  }
  /* Free mruby context */
  mrbc_context_free( mrb, c );
  /* Free interpreter instance */
  mrb_close( mrb );
}
